---
title: "Plotting and Colors in R"
author: "Chidi Justice"
date: "8/2/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Aside from the default color schemes in R, there are much better ways of handling colors in R and we should make use of them as much as possible. 
```{r}
set.seed(9)
x <- rnorm(30)
y <- rnorm(30)
plot(x, y, col = rep(1:3, each = 10), pch = 19)
legend("bottomright", legend = paste("Group", 1:3), col = 1:3, pch = 19, bty = "n")
```

This works because the default colors in R are 1 for red, 2 for blue, and 3 for red. We usually add colors to a plot, not to improve its artistic qualities, but to add another dimension to the visualisation, therefore the range and pallette of colors we use should depend on the kind of data.  R has several utilities for dealing with colors and color pallettes. `grDevices` package has two functions   
* colorRamp: Takes a pallette of colors and returns a function that takes a value between 0 and 1 indicating the extremes of the pallette (i.e it returns a function that acts like `?gray`)  
* colorRampPallette: Takes a pallette of colors and returns a function that takes integer arguments and returns a vector of colors of the size of the integer arguments. i.e it returns a function that acts like `heat.colors(n)` or `topo.colors(n)`.  

We can see the colors we can use in base R using the `colors()` function.  

##### colorRamp()  
```{r}
pal <- colorRamp(c("Red", "Blue"))
pal(0:5 / 7) # input has to be between 0 and 1
pal(0.9)
```

Notice that `pal` above is a function itself, and we can pass in a single number or a vector of numbers between 0 and 1 and it'll generate colors based on that pallette for the size of its input.  

##### colorRampPallette()  
```{r}
pal <- colorRampPalette(c("blue", "green"))
pal(2) # input indicates the size of colors in pallette we want
pal(7)
```

### RcolorBrewer package  
Both `colorRamp` and `colorRampPallette` are great if we start with a good color pallette, and this is what the **RcolorBrewer** package provides us, a good set of color pallettes which we can pass in to `colorRamp` or `colorRampPallette`.  
This package offers 3 kinds of pallettes;  
* Sequential: For numerical data that is ordered  
* Diverging: For numerical data that can be positive or negative  
* Qualitative/categorical: For qualitative unordered data  

```{r}
library(RColorBrewer)
display.brewer.all()
```

The only real function in this package is `brewer.pal`, which has two arguments ("name" and "n"), the name is the name of the pallette we want to use from those available in the package, and n is the number of colors we want from the pallette.  
```{r}
colors <- brewer.pal(3, "BuGn")
pal <- colorRampPalette(colors)
image(volcano, col = pal(20))
```

The `smoothScatter()` function is useful in making scatterplots of many data points, it gives us a color density representation of a scatterplot using a sequential pallette like that of RColowBrewer package  
```{r}
set.seed(9)
x <- rnorm(10000)
y <- rnorm(10000)
par(mfrow = c(1, 2))
smoothScatter(x, y)
plot(x, y)
```

Notice how nice the `smoothScatter` makes us see the density of the data points in the plot above (left), compared to the default scatterplot's (right). The RcolorBrewer pallette used by this function is the "Blues" pallette.  

We can also add transparency (alpha) by setting the level in the `rgb` function to get the color with its alpha. We can do it for individual colors or for a pallette  
```{r}
cols <- brewer.pal(5, "Reds")

getColorRampAlpha <- function(n, ..., alpha = 1) {
  colorRamp = colorRampPalette(...)(n)
  paste(colorRamp, sprintf("%x", ceiling(255 * alpha)), sep="")
}

image(volcano, col = getColorRampAlpha(n = 20, cols))

# The above was to add alpha to color ramp pallette. For individual colors, I can do
colour = rgb(0, 0, 0, 0.15) # the last argument being the alpha
plot(x, y, pch = 19, col = colour)
```

So from the above, we can see that we have several options of rich colors depending on the kind of data we want to plot and what we want to show. The RColorBrewer package is a great package that helps us get the right pallette for either `colorRamp` or `colorRampPallette`. Transparency can be useful to clarify plots with many points. The choice of colors can improve the readability of our plot, so always keep this in mind. Make using RColorBrewer and the base functions a habit instead of just using base colors randomly. To improve the initial plot used at the beginning  
```{r}
set.seed(9)
## Used brewer.pal.info to find the pallette I'll use
cols <- brewer.pal(n = 3, "Dark2")
pal <- colorRampPalette(cols)
col_pal <- pal(3)

x1 <- rnorm(30)
y1 <- rnorm(30)
df <- data.frame(x1, y1)
plot(x1, y1, col = rep(col_pal, each = 10), pch = 19)
legend("topright", legend = paste("Group", 1:3), col = col_pal, pch = 19, bty = "n")
```























